Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-02-09 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

Determining the early-onset flow instability is critical to accurately simulate formation, growth and coalescence of fingers, and therefore to: (a) predict displacement (sweep) efficiency and (b) design strategies to either mitigate or avoid fingering and preferential flow pathways. Mesh resolution is thus crucial to investigate Saffmann-Taylor instabilities, and adaptive dynamic mesh technology has proved to be an efficient tool to capture initial interface perturbations. Numerical simulations were performed with VR=10 using fixed and adaptive \PN[1]{2} element-pairs. Results (Figs.~\ref{fig:5regions_plottimenodeselements} and~\ref{fig:5regions_AdaptZoom}) demonstrated that by dynamically adapting the mesh to follow saturation gradients (\ie fluids interfaces), early-onset flow instability can be readily captured with further formation of long multi-fluid fingers which have themselves split into sub-fingers (see Figs.~\ref{fig:5regions_fixedmesh}i and~\ref{fig:5regions_adaptmesh}i). Overall computational cost of the simulation conducted with dynamically adapting the mesh were significantly smaller than the simulation conducted with fixed mesh (Figs.~\ref{fig:5regions_plottimenodeselements}).

\medskip
By definition the instability rises due to viscous fingering depending on capillary number ($N_{c}$) but is also a direct fuction of viscocity ratio (\ie VR). If the mobility ratio is significantly less than $5$, it is expected to find capillary fingering at relatively low flow rate and viscous fingering at relatively high flow rate. On the other hand, when the viscosity ratio is greater than $5$, the two basic forms of immiscible displacement are capillary fingering at relatively low flow rate and stable displacement at relatively high flow rate. Therefore for an \ie water/oil system, the first condition applies as the viscosity of injected water is less than the viscosity of displaced oil. Hence, the displacement front could be either capillary fingering at relatively low injection rates or viscous fingering at higher injection rates. The method that has been described in this paper as well as the simulations performed have a number of distinct advantages; (1.) The geological model can be designed from the beginning without any previous reference or geometrical limitations due to both the flexibility of method and because the geological characteristic and parameters can be easily introduced without the need of upscaling while, (2.) computational cost can be reduced, because the complex geometries and surfaces are less expensive to generate (unstructured grid in space and time) and manage using fewer grid cells. The basis of the this argument is the flexibility and variance of the adaptive and unstructured grid in time and space.

@@

\red{(3)(optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.}

\section{Acknowledgements}
Mr William Rad\"unz would like to acknowledge the support from the Brazilian Research Council (CNPq) under the \textit{Science without Borders scholarship programme}. Mr Konstantinos Christou would like to acknowledge the support of the University of Aberdeen - College of Physical Science as well as the Aberdeen Formation Evaluation Society (\textit{AFES} is an SPWLA chapter).

>>>>

Determining the early-onset flow instability is critical to accurately simulate formation, growth and coalescence of fingers, and therefore to: (a) predict displacement (sweep) efficiency and (b) design strategies to either mitigate or avoid fingering and preferential flow pathways. Mesh resolution is thus crucial to investigate Saffmann-Taylor instabilities, and adaptive dynamic mesh technology has proved to be an efficient tool to capture initial interface perturbations. Numerical simulations were performed with VR=10 using fixed and adaptive \PN[1]{2} element-pairs. Results (Figs.~\ref{fig:5regions_plottimenodeselements} and~\ref{fig:5regions_AdaptZoom}) demonstrated that by dynamically adapting the mesh to follow saturation gradients (\ie fluids interfaces), early-onset flow instability and further formation of long multi-fluid fingers, which have themselves split into sub-fingers, can be readily captured (see Figs.~\ref{fig:5regions_fixedmesh}i and~\ref{fig:5regions_adaptmesh}i). Overall computational cost of the simulation conducted with dynamically adapting the mesh were significantly smaller than the simulation conducted with fixed mesh as indicated by the number of elements/nodes (Figs.~\ref{fig:5regions_plottimenodeselements}).

Investigation of preferential flow pathways was the focus of Section~\ref{section:results_3D}, where 3D numerical simulations of multi-fluid flow through semi-pervious/impervious channelised domain were performed. Numerical data indicated preferential flow through regions of larger permeability leaving adjacent region (with low permeability) with nearly stagnant fluids (Figs.~\ref{fig:3DChannel_sat}-~\ref{fig:3DChannel_planes}). Order 0f the element-pair and mesh resolution play key-roles to capture directional flow pathways and fingers in 3D simulations.
\medskip

% (3)(optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.

\begin{comment}
\medskip
By definition the instability rises due to viscous fingering depending on capillary number ($N_{c}$) but is also a direct fuction of viscocity ratio (\ie VR). If the mobility ratio is significantly less than $5$, it is expected to find capillary fingering at relatively low flow rate and viscous fingering at relatively high flow rate. On the other hand, when the viscosity ratio is greater than $5$, the two basic forms of immiscible displacement are capillary fingering at relatively low flow rate and stable displacement at relatively high flow rate. Therefore for an \ie water/oil system, the first condition applies as the viscosity of injected water is less than the viscosity of displaced oil. Hence, the displacement front could be either capillary fingering at relatively low injection rates or viscous fingering at higher injection rates. The method that has been described in this paper as well as the simulations performed have a number of distinct advantages; (1.) The geological model can be designed from the beginning without any previous reference or geometrical limitations due to both the flexibility of method and because the geological characteristic and parameters can be easily introduced without the need of upscaling while, (2.) computational cost can be reduced, because the complex geometries and surfaces are less expensive to generate (unstructured grid in space and time) and manage using fewer grid cells. The basis of the this argument is the flexibility and variance of the adaptive and unstructured grid in time and space.

@@

\red{(3)(optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.}
\end{comment}
\section{Acknowledgements}
Mr William Rad\"unz would like to acknowledge the support from the Brazilian Research Council (CNPq) under the \textit{Science without Borders scholarship programme}. Mr Konstantinos Christou would like to acknowledge the support of the University of Aberdeen - College of Physical Science as well as the Aberdeen Formation Evaluation Society (\textit{AFES} is an SPWLA chapter).
